Assessing primitives performance\\ on multi-stage execution

### 2018-05-18 paper.tex
update table2 implem names
fix multiple things

\def\legBase{Baseline}
\def\legSmartSyntaxPlug{Syntax}
\def\legSlangPlug{Plugin}
\def\legSlang{Slang}
\def\legRTL{Slang+RTL}
\def\plotHeight{7cm}

@@

@@

We distinguish four implementations: the Baseline in pure Smalltalk, the Plugin version, the Slang version and the Slang+RTL version.

@@

\paragraph{\legSlangPlug.}
The VM plugin implementation can be easily added and removed as explained in Section \ref{sec:VMCompil}. Written in Slang, it can be compiled to C independently to the VM and is then compiled to native code through the C compiler as internal or external plugin. Some accesses to the string objects (for example, checking if they are byte strings) are done indirectly through the main VM API and these calls generate some overhead. %\sd{do we need the next sentence?}This primitive follows the old specification, the full code is available in Appendix~\ref{PluginAppendix}.

\paragraph{\legSlang.}
The Slang implementation is written inside the interpreter and has access to the full interpreter APIs. This primitive takes with two or three arguments and needs to deal with both cases in a way that the common case is efficient. The primitive is now part of the core of the VM, it cannot be easily added or removed as plugin primitives, but it is simpler to read and write. %This primitive follows the new specification, the full code is available in Appendix \ref{SlangAppendix}.

\paragraph{\legRTL.} 
In addition to the previous implementation, we've re-implemented here the primitive on top of Cog's JIT back-end. The primitive is written in Cog's RTL, an abstract assembly whose instructions compile almost one-to-one to native instructions. This primitive requires a more important amount of work to be written: the number of line of code is larger, the implementor has to be careful to generate efficient machine code (there is no C compiler to optimize the code) and debugging such code requires the use of a processor simulator such as Bochs. Since only performance critical implementations matter in this context, we've only re-implemented the case where there are two parameters. For the three parameter primitive, the VM falls back to the Slang version. %This primitive follows the new specification, the full code is available in Appendix \ref{RTLAppendix}.

@@

\vfill\null %Title not at bottom of page please

@@

\todo{Make a  pass cba}

The SlangPlugin, PureSlang and Slang+JIT implementations of this string comparison primitive have been recently implemented in the Cog VM. In this section, we compare execution speed of these primitives against the other existing implementations. We measure their respective execution times for different string sizes.% \sd{mention microbenchmarks here.}\sd{why there is only 4.3 and not something else? Because this is strange from a shape structure.}

@@

Each benchmark was measured 30 times.

@@

   \end{subfigure}
   \subcaption*{\textbf{LLVM-Mac}\vspace{-0.2cm}} %change vspace based on fig position
    \end{subfigure}
   \caption{String comparison benchmarks for strings of different sizes on different architectures.}
   \label{fig:bench}
\end{figure*}
\begin{table*}
\centering
\begin{tabular}{|c|c|c|c|c|c|c|}
  \hline
    & & 0 & 1 & 5 & 100 & 1000 \\
   \hline
  \multirow{2}{*}{SlangPlugin} &GCC& 0.4222 +- 0.0008 & 0.7674 +- 0.0011 & 1.8906 +- 0.0119 & 13.1218 +- 0.0171  & 25.3595 +- 0.1742 \\
  							   &LLVM& 0.3856 +- 0.0004 & 0.6685 +- 0.0010 & 1.6387 +- 0.0101 & 11.6376 +- 0.0576  & 23.9579 +- 0.1390 \\
   \hline
  \multirow{2}{*}{PureSlang} &GCC& 1.0288 +- 0.0046 & 1.8431 +- 0.0116 & 4.0412 +- 0.0611 & 17.7084 +- 0.1059 & 25.4258 +- 0.1621\\
 							 &LLVM& 0.2088 +- 0.0002 & 0.3649 +- 0.0064 & 0.0006 +- 0.0056 & 6.1978 +- 0.0180  & 10.6698 +- 0.0066 \\
   \hline
  \multirow{2}{*}{Slang+JIT} &GCC& 2.4290 +- 0.0029 & 3.5170 +- 0.0197  & 7.1216 +- 0.0105 & 18.5537 +- 0.0238 & 26.7626 +- 0.1861\\
 							&LLVM & 2.5718 +- 0.0035 & 3.5919 +- 0.0076 & 6.9569 +- 0.0416 & 17.2747 +- 0.1071  & 25.3991 +- 0.1811 \\
  \hline
\end{tabular} 
\vspace{0.2cm}
\caption{Benchmark results with standard errors. Average relative speed-up regarding baseline.}
\label{tab:detail_bench}
\end{table*}

@@

In each benchmark, we run a string comparison in a loop and measure the total execution time.
The number of iterations of the loop depends on the string size: the comparison is repeated between 1414 and 10000 times for a short string (5 characters or less), and between 100 and 316 times for a long string (100 characters or more). The warm-up time of the JIT is in this context not noticeable on the performance results: the primitive is compiled to native code usually at the second call as explained in Section \ref{sec:VMexec} and the VM sampling profiler \cite{CogProfiler} confirms no significant time is spent in the interpreter loop (Between 0 and 1 sample over dozens of thousands are shown in the interpreter loop).

The high number of iterations allows the benchmarks to run at least 1 second in the slower implementations to avoid processor and OS noises.
The loop overhead is measured and removed from the measurements.

The two compared strings are equal as this situation corresponds to the worst scenario for most implementations. %There is one less branch in the two plugin implementations in this case.

\subsection{Results} 
We performed the benchmarks upon five different string sizes: 0, 1, 5, 100 and 1000 characters, using two different compilers: GCC on one hand and LLVM on the other hand. Figure~\ref{fig:bench} shows the results in terms of relative speed-up towards the baseline implementation, respectively for GCC and LLVM.

We can draw the following patterns from these results:
\begin{itemize}
\item The execution time of the Plugin implementations is greater than the baseline's for empty strings (0,58 times slower) or 1-character strings (0,23 times slower).
\item The Slang+JIT implementation is more efficient than the baseline in every case. Nevertheless, the performance gap between this version and the Plugin and Slang ones tends to decrease as the string size increases, especially for GCC. For 1000-characters strings, the Slang+JIT version goes 0,25 times faster than the Plugin and the Slang versions.
\item The compiler used (either GCC or LLVM) does lead to significant differences in terms of performance for the Slang implementation. 
\end{itemize}

@@

\paragraph{Stack switch overhead.}
As mentioned in Section~\ref{sec:VMexec}, all implementations written in Slang come with an overhead due to the switch between the C and Smalltalk stacks. This overhead is clearly visible in the bench results in Figure~\ref{fig:bench} for small strings (size 1 or less): the Plugin implementation is between 57,8 (gcc) to 61,4\% (llvm) slower than the baseline's for empty strings.
As the string size goes larger, the overhead is less significant because it is absorbed by other performance gains.
Regarding this criteria, the Slang+JIT version is always the fastest as it is not impacted by the switch overhead.

\paragraph{Optimization potential on large strings comparison.}
The Plugin come with an overhead due the use of an interface. Besides, cross-file inlining is not performed for plugins, like said in Section~\ref{sec:implem_cmp}. These are some of the factors explaining the performance gap between these versions and the Slang and Slang+JIT ones.
When it comes to comparing large strings (100 characters and more), the execution time is mainly spent in the comparison loop (Figure~\ref{fig:Asm}). However, the switch overhead is still noticeable in the Slang implementation when strings of 100 characters are compared: Slang is 17.7 times faster than Baseline while Slang+JIT is 18,6 times faster than Baseline. 
This last observation only applies to the GCC results: the code generated by the LLVM compiler induces a longer execution time, as explained in the next paragraph.

@@

In this section, we compared the performance of the different implementations against a baseline (\ie{pure Smalltalk code}) implementation. The results showed the Slang+JIT implementation was always the fastest. Otherwise, most of the performance gaps between the primitive versions could be explained by 3 criteria: stack swithing, optimization for large strings, and the control over generated code.

>>>>

\def\legBase{Baseline\xspace}
\def\legSmartSyntaxPlug{Syntax\xspace}
\def\legSlangPlug{Plugin\xspace}
\def\legSlang{Slang\xspace}
\def\legRTL{Slang+RTL\xspace}
\def\plotHeight{7cm}

@@

Conceptually, the primitive with two parameters first checks if the operands are ByteStrings and fails if they are not. It then extract the ByteString sizes from their object header and computes the minimum size. Lastly, it iterates over the two ByteStrings until the minimum size is reached. If a difference is found, if answers the difference between the two different characters. If no difference is found, it answers the difference between the two string sizes.

@@

We distinguish four implementations: the \legBase in pure Smalltalk, the \legSlangPlug version, the \legSlang version and the \legRTL version.

@@

\paragraph{\legSlangPlug.}
Written in Slang, the VM plugin implementation is compiled to C independently to the VM and is then compiled to native code through the C compiler as an internal plugin, without enabling linking time optimizations. Accesses to the ByteString object header (to check if they are actually ByteString and not other objects as well as extracting their size), and accesses to the stack (to read the operands) are done indirectly through the main VM API since they're dependent on the memory representation of objects and stack layout used. These calls generate some overhead since they cannot be inlined. To avoid such calls in the main comparison loop, the primitive requests the main VM to provide a pointer to the first byte of each string and then assumes all bytes are then contiguous.

\paragraph{\legSlang.}
The Slang implementation is written inside the interpreter and has access to the full interpreter APIs. It is compiled to C and then to native code as part of the main VM. There is no overhead in accessing the object memory representation or the stack as in the \legSlangPlug version.

\paragraph{\legRTL.} 
In addition to the previous implementation, we've re-implemented here the primitive on top of Cog's JIT back-end. The primitive is written in Cog's RTL, an abstract assembly whose instructions compile almost one-to-one to native instructions. This primitive requires a more important amount of work to be written: we had to be careful to generate efficient machine code (there is no C compiler to optimize the code) and debugging such code requires the use of a processor simulator such as Bochs. Since only performance critical implementations matter in this context, we've only re-implemented the case where there are two parameters. For the three parameter primitive, the VM falls back to the Slang version.

@@

%\vfill\null %Title not at bottom of page please

@@

The \legSlangPlug, \legSlang and \legRTL implementations of this string comparison primitive have been implemented in the Cog VM. In this section, we compare execution speed of these primitives against the other existing implementations. We measure their respective execution times for different string sizes.% \sd{mention microbenchmarks here.}\sd{why there is only 4.3 and not something else? Because this is strange from a shape structure.}

@@

@@

   \end{subfigure}
   \subcaption*{\textbf{LLVM-Mac}\vspace{-0.2cm}} %change vspace based on fig position
    \end{subfigure}
   \caption{String comparison benchmarks for strings of different sizes on different architectures.}
   \label{fig:bench}
\end{figure*}
\begin{table*}
\centering
\begin{tabular}{|c|c|c|c|c|c|c|}
  \hline
    & & 0 & 1 & 5 & 100 & 1000 \\
\hline
  \multirow{2}{*}{\legSlangPlug} &GCC-Linux& 0.4222 $\pm$ 0.0008 & 0.7674 $\pm$ 0.0011 & 1.8906 $\pm$ 0.0119 & 13.1218 $\pm$ 0.0171  & 25.3595 $\pm$ 0.1742 \\
  						&LLVM-Mac& 0.3856 $\pm$ 0.0004 & 0.6685 $\pm$ 0.0010 & 1.6387 $\pm$ 0.0101 & 11.6376 $\pm$ 0.0576  & 23.9579 $\pm$ 0.1390 \\
   \hline
  \multirow{2}{*}{\legSlang} &GCC-Linux& 1.0288 $\pm$ 0.0046 & 1.8431 $\pm$ 0.0116 & 4.0412 $\pm$ 0.0611 & 17.7084 $\pm$ 0.1059 & 25.4258 $\pm$ 0.1621\\
 					 &LLVM-Mac& 0.2088 $\pm$ 0.0002 & 0.3649 $\pm$ 0.0064 & 0.8953 $\pm$ 0.0056 & 6.1978 $\pm$ 0.0180  & 10.6698 $\pm$ 0.0066 \\
   \hline
  \multirow{2}{*}{\legRTL}  &GCC-Linux& 2.4290 $\pm$ 0.0029 & 3.5170 $\pm$ 0.0197  & 7.1216 $\pm$ 0.0105 & 18.5537 $\pm$ 0.0238 & 26.7626 $\pm$ 0.1861\\
 				 	&LLVM-Mac & 2.5718 $\pm$ 0.0035 & 3.5919 $\pm$ 0.0076 & 6.9569 $\pm$ 0.0416 & 17.2747 $\pm$ 0.1071  & 25.3991 $\pm$ 0.1811 \\
  \hline
\end{tabular} 
\vspace{0.2cm}
\caption{Benchmark results with standard errors. Average relative speed-up regarding baseline.}
\label{tab:detail_bench}
\end{table*}

@@

Each benchmark was measured 30 times. Each time, we run a string comparison in a loop and measure the total execution time.

The number of iterations of the loop depends on the string size: the comparison is repeated between 1414 and 10000 times for a short string (5 characters or less), and between 100 and 316 times for a long string (100 characters or more). The warm-up time of the JIT is in this context not noticeable on the performance results: the primitive is compiled to native code usually at the second call as explained in Section \ref{sec:VMexec} and the VM sampling profiler \cite{CogProfiler} confirms no significant time is spent in the interpreter loop (Between 0 and 1 sample over dozens of thousands are shown in the interpreter loop). The number of iterations allows the benchmarks to run for at least 1 second in the slower implementations to avoid processor and OS noises.

The loop overhead is measured and removed from the measurements. The two compared strings are equal as this situation corresponds to the worst scenario for most implementations. %There is one less branch in the two plugin implementations in this case.

\subsection{Results} 
We performed the benchmarks upon five different string sizes: 0, 1, 5, 100 and 1000 characters. In the case of small strings, most of the time is spent activating the primitive and in the code before the main comparison loop, the primitive is performed with a little overall number of native instructions. In the case of large strings, the time spent executing the primitive is dominated by the main comparison loop. Figure~\ref{fig:bench} shows the results in terms of relative speed-up towards the baseline implementation, respectively for GCC-Linux and LLVM-Mac.

We can draw the following patterns from these results:
\begin{itemize}
\item The \legSlangPlug implementation is slower than the baseline version for small strings (0,58 times slower for empty strings, 0,23 times slower for 1-character strings). This is due to the overhead of the calls to the main VM to access the object headers and the stack. On large strings, the performance is comparable to 
\item The \legRTL implementation is more efficient than the baseline in every case. Nevertheless, the performance gap between this version and the \legSlangPlug and \legSlang ones tends to decrease as the string size increases, especially for GCC. For 1000-characters strings, the \legRTL version goes 0,25 times faster than the \legSlangPlug and the \legSlang versions.
\item The compiler used (either GCC or LLVM) does lead to significant differences in terms of performance for the Slang implementation. 
\end{itemize}

@@

\paragraph{Stack switch overhead.}
As mentioned in Section~\ref{sec:VMexec}, all implementations written in Slang come with an overhead due to the switch between the C and Smalltalk stacks. This overhead is clearly visible in the bench results in Figure~\ref{fig:bench} for small strings (size 1 or less): the Plugin implementation is between 57,8 (gcc) to 61,4\% (llvm) slower than the baseline's for empty strings.
As the string size goes larger, the overhead is less significant because it is absorbed by other performance gains.
Regarding this criteria, \legRTL version is always the fastest as it is not impacted by the switch overhead.

\paragraph{Optimization potential on large strings comparison.}
The Plugin come with an overhead due the use of an interface. Besides, cross-file inlining is not performed for plugin, like said in Section~\ref{sec:implem_cmp}. These are some of the factors explaining the performance gap between these versions and the \legSlang and \legRTL ones.
When it comes to comparing large strings (100 characters and more), the execution time is mainly spent in the comparison loop (Figure~\ref{fig:Asm}). However, the switch overhead is still noticeable in the Slang implementation when strings of 100 characters are compared: Slang is 17.7 times faster than Baseline while Slang+JIT is 18,6 times faster than Baseline. 
This last observation only applies to the GCC results: the code generated by the LLVM compiler induces a longer execution time, as explained in the next paragraph.

@@

In this section, we compared the performance of the different implementations against a baseline (\emph{i.e.,}{pure Smalltalk code}) implementation. The results showed the Slang+JIT implementation was always the fastest. Otherwise, most of the performance gaps between the primitive versions could be explained by 3 criteria: stack swithing, optimization for large strings, and the control over generated code.
